---
title: "Multinichenet_interpretation"
output: html_document
date: "2024-12-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
library(Seurat)
library(Matrix)
library(nichenetr) #install_github("saeyslab/nichenetr")
library(multinichenetr) #install_github("saeyslab/multinichenetr")
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
```

```{r}
multinichenet_output <- readRDS("~/palmer_scratch/multinichenet_output.rds")
```

```{r}
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  rank_per_group = FALSE
  )
```

```{r}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{r}
group_oi = "normal"

prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 50, 
  rank_per_group = TRUE
  )

prioritized_tbl_oi_normal_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi) 
```

```{r}
prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_normal_50$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_normal_50)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), prioritized_tbl_oi$receiver %>% unique()) %>% sort()

colors_sender = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)
colors_receiver = RColorBrewer::brewer.pal(n = length(senders_receivers), name = 'Spectral') %>% magrittr::set_names(senders_receivers)

circos_list = make_circos_group_comparison(prioritized_tbl_oi, colors_sender, colors_receiver)
```

```{}
```

```{}
```

### Interpretable bubble plots

```{r}
#for nash group
group_oi = "nash"

prioritized_tbl_oi_nash_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r, fig.width=15, fig.height=12}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_nash_50 %>% inner_join(lr_network_all)
  )
plot_oi
```
```{r}
#for normal group
group_oi = "normal"

prioritized_tbl_oi_normal_50 = prioritized_tbl_oi_all %>% 
  filter(group == group_oi)
```

```{r,fig.width=15, fig.height=12}
plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_normal_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

Typically, there are way more than 50 differentially expressed and active ligand-receptor pairs per group across all sender-receiver combinations. Therefore it might be useful to zoom in on specific cell types as senders/receivers:

```{r,fig.width=15, fig.height=12}
#kupffer cells as receiver
group_oi = "nash"
prioritized_tbl_oi_nash_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  50, 
  groups_oi = group_oi, 
  receivers_oi = "Kupffer.cells"
  ) 

plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_nash_50 %>% inner_join(lr_network_all)
  )
plot_oi
```
```{r,fig.width=15, fig.height=12}
#kupffer cells as sender
group_oi = "nash"
prioritized_tbl_oi_nash_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  50, 
  groups_oi = group_oi, 
  senders_oi = "Kupffer.cells"
  ) 

plot_oi = make_sample_lr_prod_activity_plots_Omnipath(
  multinichenet_output$prioritization_tables, 
  prioritized_tbl_oi_nash_50 %>% inner_join(lr_network_all)
  )
plot_oi
```

## Visualize sender-agnostic ligand activities for each receiver-group combination
```{r, fig.width=15, fig.height=14}
ligands_oi = multinichenet_output$prioritization_tables$ligand_activities_target_de_tbl %>% 
  inner_join(contrast_tbl) %>% 
  group_by(group, receiver) %>% filter(direction_regulation == "up") %>% 
  distinct(ligand, receiver, group, activity) %>% 
  top_n(5, activity) %>% 
  pull(ligand) %>% unique()

plot_oi = make_ligand_activity_plots(
  multinichenet_output$prioritization_tables, 
  ligands_oi, 
  contrast_tbl,
  widths = NULL)
plot_oi
```

### Visualization of ligand-to-target signaling paths
```{r}
if(organism == "human"){
  sig_network = readRDS(url("https://zenodo.org/record/7074291/files/signaling_network_human_21122021.rds")) %>% 
    mutate(from = make.names(from), to = make.names(to))
  
  gr_network = readRDS(url("https://zenodo.org/record/7074291/files/gr_network_human_21122021.rds")) %>% 
    mutate(from = make.names(from), to = make.names(to))
  
  ligand_tf_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_tf_matrix_nsga2r_final.rds"))
  colnames(ligand_tf_matrix) = colnames(ligand_tf_matrix) %>% make.names()
  rownames(ligand_tf_matrix) = rownames(ligand_tf_matrix) %>% make.names()
  
  weighted_networks = readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final.rds"))
  weighted_networks$lr_sig = weighted_networks$lr_sig %>% mutate(from = make.names(from), to = make.names(to))
  weighted_networks$gr = weighted_networks$gr %>% mutate(from = make.names(from), to = make.names(to))
  
} else if(organism == "mouse"){
  sig_network = readRDS(url("https://zenodo.org/record/7074291/files/signaling_network_mouse_21122021.rds")) %>% 
    mutate(from = make.names(from), to = make.names(to))
  
  gr_network = readRDS(url("https://zenodo.org/record/7074291/files/gr_network_mouse_21122021.rds")) %>% 
    mutate(from = make.names(from), to = make.names(to))
  
  ligand_tf_matrix = readRDS(url("https://zenodo.org/record/7074291/files/ligand_tf_matrix_nsga2r_final_mouse.rds"))
  colnames(ligand_tf_matrix) = colnames(ligand_tf_matrix) %>% make.names()
  rownames(ligand_tf_matrix) = rownames(ligand_tf_matrix) %>% make.names()
  
  weighted_networks = readRDS(url("https://zenodo.org/record/7074291/files/weighted_networks_nsga2r_final_mouse.rds"))
  weighted_networks$lr_sig = weighted_networks$lr_sig %>% mutate(from = make.names(from), to = make.names(to))
  weighted_networks$gr = weighted_networks$gr %>% mutate(from = make.names(from), to = make.names(to))
}
```
Define which ligand and target genes you want to focus on
```{r}
network$links %>% filter(sender_ligand == "Kupffer.cells_Ccl3" & direction_regulation == "up" & group == "nash")
```
```{r}
ligand_oi = "Ccl3"
receptor_oi = "Ccl3"
targets_all = c("Ccl3", "Ccl3")
  
active_signaling_network = nichenetr::get_ligand_signaling_path_with_receptor(
  ligand_tf_matrix = ligand_tf_matrix, 
  ligands_all = ligand_oi, 
  receptors_all = receptor_oi, 
  targets_all = targets_all, 
  weighted_networks = weighted_networks, 
  top_n_regulators = 3
  )

data_source_network = nichenetr::infer_supporting_datasources(
  signaling_graph_list = active_signaling_network,
  lr_network = lr_network %>% dplyr::rename(from = ligand, to = receptor), 
  sig_network = sig_network, 
  gr_network = gr_network
  )
```

```{r, fig.width=15}
active_signaling_network_min_max = active_signaling_network
active_signaling_network_min_max$sig = active_signaling_network_min_max$sig %>% mutate(weight = ((weight-min(weight))/(max(weight)-min(weight))) + 0.75)
active_signaling_network_min_max$gr = active_signaling_network_min_max$gr %>% mutate(weight = ((weight-min(weight))/(max(weight)-min(weight))) + 0.75)
colors = c("ligand" = "purple", "receptor" = "orange", "target" = "royalblue", "mediator" = "grey60")
#ggraph_signaling_path = suppressWarnings(make_ggraph_signaling_path(active_signaling_network_min_max, colors, ligand_oi, receptor_oi, targets_all))
ggraph_signaling_path = make_ggraph_signaling_path(
  active_signaling_network_min_max, 
  colors, 
  ligand_oi, 
  receptor_oi, 
  targets_all)
ggraph_signaling_path$plot
```

data_source_network %>% head()
