---
title: "multinichenet"
output: html_document
date: "2024-12-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(Seurat)
library(Matrix)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(nichenetr)
library(multinichenetr)
```

```{r}
dir <- '/home/cbb575_nb752/palmer_scratch/R_data'
counts <- readMM(paste0(dir,'/counts.mtx'))
dim(counts)
```

```{r}
cellMeta <- read.csv(paste0(dir,'/counts_cellMeta.csv'))
geneMeta <- read.csv(paste0(dir,'/counts_geneMeta.csv'))
```

```{r}
head(cellMeta)
dim(cellMeta)
```

```{r}
dim(geneMeta)
head(geneMeta)
```

```{r}
anyDuplicated(cellMeta$Barcode)
cellMeta$Barcode <- make.unique(cellMeta$Barcode)
```

```{r}
rownames(counts) <- cellMeta$Barcode
colnames(counts) <- geneMeta$GeneName

seo <- CreateSeuratObject(counts = t(counts))
### Set the meta data
seo@meta.data<-cbind(cellMeta,seo@meta.data)
rownames(seo@meta.data)<-colnames(seo)
```

```{r}
bootstrap_pseudo_samples <- function(seo, treatment_col = "treatment", n_pseudo = 2) {
  # Extract metadata
  meta <- seo@meta.data
  
  # Initialize a list to store bootstrapped rows
  bootstrapped_meta <- list()
  
  # Iterate over treatment groups
  for (treatment in unique(meta[[treatment_col]])) {
    # Subset cells for the current treatment
    cells <- rownames(meta[meta[[treatment_col]] == treatment, ])
    
    # Bootstrap cells to create `n_pseudo` pseudo-samples
    pseudo_assignments <- rep(1:n_pseudo, ceiling(length(cells) / n_pseudo))[1:length(cells)]
    pseudo_assignments <- sample(pseudo_assignments)  # Randomize assignments
    
    # Create new metadata for bootstrapped pseudo-samples
    for (pseudo in 1:n_pseudo) {
      bootstrapped_cells <- cells[pseudo_assignments == pseudo]
      pseudo_sample_meta <- meta[bootstrapped_cells, ]
      pseudo_sample_meta$pseudo_sample <- paste0(treatment, "_pseudo_", pseudo)
      bootstrapped_meta[[length(bootstrapped_meta) + 1]] <- pseudo_sample_meta
    }
  }
  
  # Combine all bootstrapped metadata
  new_meta <- do.call(rbind, bootstrapped_meta)
  
  # Update the Seurat object
  seo@meta.data <- new_meta
  seo <- subset(seo, cells = rownames(new_meta))  # Subset Seurat object to new metadata
  
  return(seo)
}


n_pseudo <- 2
seo <- bootstrap_pseudo_samples(seo, treatment_col = "treatment", n_pseudo = n_pseudo)
```


```{r}
### Save the seurat object
saveRDS(seo, file = "seo.rds")
```

```{r}
seo <- readRDS("seo.rds")
```

# Load NicheNet's ligand-receptor network and ligand-target matrix

```{r}
organism = "mouse"
options(timeout = 120)

if(organism == "human"){
  
  lr_network_all = 
    readRDS(url(
      "https://zenodo.org/record/10229222/files/lr_network_human_allInfo_30112033.rds"
      )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
} else if(organism == "mouse"){
  
  lr_network_all = readRDS(url(
    "https://zenodo.org/record/10229222/files/lr_network_mouse_allInfo_30112033.rds"
    )) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
  ligand_target_matrix = readRDS(url(
    "https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final_mouse.rds"
    ))
  
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
  
}
```


## Read in SingleCellExperiment Object

```{r}
sce = Seurat::as.SingleCellExperiment(seo, assay = "RNA")
sce = alias_to_symbol_SCE(sce, "mouse") %>% makenames_SCE()
```


# Prepare the settings of the MultiNicheNet cell-cell communication analysis

## Define in which metadata columns we can find the group, sample and cell type IDs

```{r}
sample_id = "pseudo_sample"
group_id = "treatment"
celltype_id = "cell_type"

covariates = NA
batches = NA
```

```{r}
SummarizedExperiment::colData(sce)[, celltype_id] <- 
    make.names(SummarizedExperiment::colData(sce)[, celltype_id])

SummarizedExperiment::colData(sce)[, sample_id] <- 
    make.names(SummarizedExperiment::colData(sce)[, sample_id])

SummarizedExperiment::colData(sce)[, group_id] <- 
    make.names(SummarizedExperiment::colData(sce)[, group_id])
```

## Define the contrasts of interest.

```{r}
contrasts_oi = c("'normal-nash','nash-normal'") 
contrast_tbl = tibble(contrast = c("normal-nash","nash-normal"), group = c("normal","nash"))
```

## Define the sender and receiver cell types of interest.

```{r}
senders_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
receivers_oi = SummarizedExperiment::colData(sce)[,celltype_id] %>% unique()
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]
```

```{r}
conditions_keep = c("normal","nash")
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep
          ]
```

## Running the MultiNicheNet core analysis

### Cell-type filtering: determine which cell types are sufficiently present


```{r}
min_cells = 10
abundance_info = get_abundance_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  min_cells = min_cells, 
  senders_oi = senders_oi, receivers_oi = receivers_oi, 
  batches = batches
  )
```

```{R,fig.width=12, fig.height=10}
abundance_info$abund_plot_sample
```

### Cell type filtering based on cell type abundance information

```{r}
abundance_df_summarized = abundance_info$abundance_data %>% 
  mutate(keep = as.logical(keep)) %>% 
  group_by(group_id, celltype_id) %>% 
  summarise(samples_present = sum((keep)))

celltypes_absent_one_condition = abundance_df_summarized %>% 
  filter(samples_present == 0) %>% pull(celltype_id) %>% unique() 
# find truly condition-specific cell types by searching for cell types 
# truely absent in at least one condition

celltypes_present_one_condition = abundance_df_summarized %>% 
  filter(samples_present >= 1) %>% pull(celltype_id) %>% unique() 
# require presence in at least 2 samples of one group so 
# it is really present in at least one condition

condition_specific_celltypes = intersect(
  celltypes_absent_one_condition, 
  celltypes_present_one_condition)

total_nr_conditions = SummarizedExperiment::colData(sce)[,group_id] %>% 
  unique() %>% length() 

absent_celltypes = abundance_df_summarized %>% 
  filter(samples_present < 1) %>% 
  group_by(celltype_id) %>% 
  count() %>% 
  filter(n == total_nr_conditions) %>% 
  pull(celltype_id)
  
print("condition-specific celltypes:")
print(condition_specific_celltypes)

print("absent celltypes:")
print(absent_celltypes)
```

```{r}
analyse_condition_specific_celltypes = FALSE
if(analyse_condition_specific_celltypes == TRUE){
  senders_oi = senders_oi %>% setdiff(absent_celltypes)
  receivers_oi = receivers_oi %>% setdiff(absent_celltypes)
} else {
  senders_oi = senders_oi %>% 
    setdiff(union(absent_celltypes, condition_specific_celltypes))
  receivers_oi = receivers_oi %>% 
    setdiff(union(absent_celltypes, condition_specific_celltypes))
}

sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)
          ]
```


### Gene filtering: determine which genes are sufficiently expressed in each present cell type

```{r}
min_sample_prop = 0.50
fraction_cutoff = 0.05

frq_list = get_frac_exprs(
  sce = sce, 
  sample_id = sample_id, celltype_id =  celltype_id, group_id = group_id, 
  batches = batches, 
  min_cells = min_cells, 
  fraction_cutoff = fraction_cutoff, min_sample_prop = min_sample_prop)
```

```{r}
#Now only keep genes that are expressed by at least one cell type:
genes_oi = frq_list$expressed_df %>% 
  filter(expressed == TRUE) %>% pull(gene) %>% unique() 
sce = sce[genes_oi, ]
```


### Pseudobulk expression calculation: determine and normalize per-sample pseudobulk expression levels for each expressed gene in each present cell type


```{r}
abundance_expression_info = process_abundance_expression_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  min_cells = min_cells, 
  senders_oi = senders_oi, receivers_oi = receivers_oi, 
  lr_network = lr_network, 
  batches = batches, 
  frq_list = frq_list, 
  abundance_info = abundance_info)
```

```{r}
abundance_expression_info$celltype_info$pb_df %>% head()
```

```{R}
abundance_expression_info$celltype_info$pb_df_group %>% head()
```


```{r}
abundance_expression_info$sender_receiver_info$pb_df %>% head()

abundance_expression_info$sender_receiver_info$pb_df_group %>% head()
```

## Differential expression (DE) analysis: determine which genes are differentially expressed

```{r}
DE_info = get_DE_info(
  sce = sce, 
  sample_id = sample_id, group_id = group_id, celltype_id = celltype_id, 
  batches = batches, covariates = covariates, 
  contrasts_oi = contrasts_oi, 
  min_cells = min_cells, 
  expressed_df = frq_list$expressed_df)
```

```{r}
DE_info$celltype_de$de_output_tidy %>% head()
```


```{r, fig.width=15, fig.height=10}
DE_info$hist_pvals
```

These distributions look fine (uniform distribution, except peak at p-value <= 0.05), so we will continue using these regular p-values. 

```{r}
empirical_pval = FALSE
if(empirical_pval == TRUE){
  DE_info_emp = get_empirical_pvals(DE_info$celltype_de$de_output_tidy)
  celltype_de = DE_info_emp$de_output_tidy_emp %>% select(-p_val, -p_adj) %>% 
    rename(p_val = p_emp, p_adj = p_adj_emp)
} else {
  celltype_de = DE_info$celltype_de$de_output_tidy
} 
```

### Combine DE information for ligand-senders and receptors-receivers

```{r}
sender_receiver_de = combine_sender_receiver_de(
  sender_de = celltype_de,
  receiver_de = celltype_de,
  senders_oi = senders_oi,
  receivers_oi = receivers_oi,
  lr_network = lr_network
)

sender_receiver_de %>% head(20)
```

## Ligand activity prediction: use the DE analysis output to predict the activity of ligands in receiver cell types and infer their potential target genes

### Assess geneset_oi-vs-background ratios for different DE output tresholds prior to the NicheNet ligand activity analysis

We will first inspect the geneset_oi-vs-background ratios for the default tresholds:

```{r}
logFC_threshold = 0.50
p_val_threshold = 0.05
p_val_adj = FALSE 


geneset_assessment = contrast_tbl$contrast %>% 
  lapply(
    process_geneset_data, 
    celltype_de, logFC_threshold, p_val_adj, p_val_threshold
  ) %>% 
  bind_rows() 
geneset_assessment
```

```{r}
geneset_assessment_adjustedPval = contrast_tbl$contrast %>% 
  lapply(
    process_geneset_data, 
    celltype_de, logFC_threshold, p_val_adj = TRUE, p_val_threshold
    ) %>% 
  bind_rows() 
geneset_assessment_adjustedPval
```


### Perform the ligand activity analysis and ligand-target inference

After the ligand activity prediction, we will also infer the predicted target genes of these ligands in each contrast. For this ligand-target inference procedure, we also need to select which top n of the predicted target genes will be considered (here: top 250 targets per ligand). This parameter will not affect the ligand activity predictions. It will only affect ligand-target visualizations and construction of the intercellular regulatory network during the downstream analysis.

```{r}
top_n_target = 250
verbose = TRUE
cores_system = 1
n.cores = min(cores_system, celltype_de$cluster_id %>% unique() %>% length()) 

ligand_activities_targets_DEgenes = suppressMessages(suppressWarnings(
  get_ligand_activities_targets_DEgenes(
    receiver_de = celltype_de,
    receivers_oi = intersect(receivers_oi, celltype_de$cluster_id %>% unique()),
    ligand_target_matrix = ligand_target_matrix,
    logFC_threshold = logFC_threshold,
    p_val_threshold = p_val_threshold,
    p_val_adj = p_val_adj,
    top_n_target = top_n_target,
    verbose = verbose, 
    n.cores = n.cores
  )
))
```

```{r}
ligand_activities_targets_DEgenes$ligand_activities %>% head(20)
```


## Prioritization: rank cell-cell communication patterns through multi-criteria prioritization

```{r}
ligand_activity_down = FALSE
sender_receiver_tbl = sender_receiver_de %>% distinct(sender, receiver)

metadata_combined = SummarizedExperiment::colData(sce) %>% tibble::as_tibble()

if(!is.na(batches)){
  grouping_tbl = metadata_combined[,c(sample_id, group_id, batches)] %>% 
    tibble::as_tibble() %>% distinct()
  colnames(grouping_tbl) = c("sample","group",batches)
} else {
  grouping_tbl = metadata_combined[,c(sample_id, group_id)] %>% 
    tibble::as_tibble() %>% distinct()
  colnames(grouping_tbl) = c("sample","group")
}

prioritization_tables = suppressMessages(generate_prioritization_tables(
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de = sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    contrast_tbl = contrast_tbl,
    sender_receiver_tbl = sender_receiver_tbl,
    grouping_tbl = grouping_tbl,
    scenario = "regular", # all prioritization criteria will be weighted equally
    fraction_cutoff = fraction_cutoff, 
    abundance_data_receiver = abundance_expression_info$abundance_data_receiver,
    abundance_data_sender = abundance_expression_info$abundance_data_sender,
    ligand_activity_down = ligand_activity_down
  ))

prioritization_tables$group_prioritization_tbl %>% head(20)
```

## Calculate the across-samples expression correlation between ligand-receptor pairs and target genes

```{r}
lr_target_prior_cor = lr_target_prior_cor_inference(
  receivers_oi = prioritization_tables$group_prioritization_tbl$receiver %>% unique(), 
  abundance_expression_info = abundance_expression_info, 
  celltype_de = celltype_de, 
  grouping_tbl = grouping_tbl, 
  prioritization_tables = prioritization_tables, 
  ligand_target_matrix = ligand_target_matrix, 
  logFC_threshold = logFC_threshold, 
  p_val_threshold = p_val_threshold, 
  p_val_adj = p_val_adj
  )
```

# Save all the output of MultiNicheNet

```{r}
path = "./"

multinichenet_output = list(
    celltype_info = abundance_expression_info$celltype_info,
    celltype_de = celltype_de,
    sender_receiver_info = abundance_expression_info$sender_receiver_info,
    sender_receiver_de =  sender_receiver_de,
    ligand_activities_targets_DEgenes = ligand_activities_targets_DEgenes,
    prioritization_tables = prioritization_tables,
    grouping_tbl = grouping_tbl,
    lr_target_prior_cor = lr_target_prior_cor
  ) 
multinichenet_output = make_lite_output(multinichenet_output)

save = TRUE
if(save == TRUE){
  saveRDS(multinichenet_output, paste0(path, "multinichenet_output.rds"))

}
```





